Addition and removal energies of circular quantum dots

### 2017-04-06 Manuscript/paper.tex
Update paper and figures

\input{head_plain.tex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{siunitx}
\usepackage{subcaption}

@@

\begin{document}
\title{Studies of quantum dots}

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{J\o rgen H\o gberget}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
\author{Sam Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{Scott Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract}
  We present and compare several many-body methods used to study a quantum dot system.  We calculate the approximate ground state energy of electrons in a circularly symmetric potential well using a harmonic oscillator basis optimized via Hartree-Fock (HF).  We further improve the ground state energy using in-medium similarity renormalization group (IM-SRG).  Additionally, we calculate the addition and removal energies using quasidegenerate perturbation theory (QDPT).  Our results are benchmarked against quasi-exact diffusion Monte Carlo (DMC) and coupled cluster (CC) calculations.
\end{abstract}

@@

Understanding the behavior of strongly confined electrons is of fundamental
interest for solving many-body problems.  Quantum dots, e.g. electrons
confined in semiconducting heterostructures, are of particular interest since
they exhibit, due to their small size, discrete quantum levels.  Under these
conditions, typical quantum phenomena like tunneling, entanglement and
magnetization can all be observed \cite{reimann2002,engel1993}.  Since quantum
dots are manufactured and designed artificially at the laboratory, their
quantum levels can be arbitrarily tuned by changing the external field or the
size and shape of the system.  As a consequence, quantum dots provide a high
level of control for the dynamics and correlation of the electrons, which
makes them perfectly suited to study quantum effects in practice.  Since their
ground state shows similar shell structures and magic numbers as seen for
atoms and nuclei \cite{tarucha1996}, these systems give the opportunity to
study electronic systems without the presence of a nucleus affecting the
electrons.  Apart from their relevance for theoretical research in quantum
physics, quantum dots offer a wide variety of applications: In particular,
their electrical and optical properties make them attractive for the use in
laser technology \cite{strauf2010,5075760} and solar cells
\cite{jenks:013111,doi:10.1021/cr900289f}, but they are also used in quantum
computers\cite{PhysRevA.57.120} and medical imaging \cite{Ben-Ari02042003}.

In order to properly understand the properties of quantum dots and make
theoretical predictions of their behavior in various applications, it is
necessary to study features like ground state energy and correlation
effects. Since there generally does not exist an analytical solution for
quantum dots except for ones containing only two electrons or with specific
values of the external field \cite{PhysRevA.48.3561}, the development of
appropriate numerical few- and many-body methods is required to solve the
problem.

Several \textit{ab initio} methods have been applied to these systems, in
particular variational and diffusion Monte Carlo (VMC and DMC)
\cite{PhysRevB.68.035304,%
  PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780}, large-scale
diagonalization/full configuration interaction (FCI)
\cite{JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} and coupled
cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,%
  PhysRevB.84.115302}.

Additionally to those approaches, another very promising first-principles
method has recently been developed.  The similarity renormalization group
(SRG) is a family of methods that drives the Hamiltonian toward a band- or
block-diagonal form through a continuous sequence of unitary transformations,
reducing the coupling between a subspace of interest -- such as the ground
state -- and the remaining Hilbert space.  It has successfully been applied to
study systems with different underlying potentials and to analyse their
binding energy and other observables, especially in nuclear theory
\cite{ScottSRG,PhysRevC.75.061001,SRGThreeDim}.  Aside from usual
\emph{free-space} SRG, where the Hamiltonian is constructed with respect to
the vacuum state with no physical particles, it is possible to evolve a
Hamiltonian via SRG \emph{in medium} through the normal-ordering technique,
which enables the Hamiltonian to be constructed with respect to some reference
state containing a finite number of particles.  In-medium SRG (IM-SRG) can
further take advantage of the normal-ordering process to account for 3-body
forces in an approximate manner without resorting to the full, computationally
intensive 3-body machinery.  The method allows considerable simplifications to
the problem and makes numerical calculations much more efficient \cite{IMSRG}.

This article is organized as follows: Section \ref{sec:formalism} introduces
first (\ref{subsec:modelHamiltonian}) the Hamiltonian we use to model circular
quantum dots, and gives afterwards an overview of the IM-SRG
(\ref{subsec:imsrgmethod}) and DMC (\ref{subsec:DMC}) formalisms.  Our results are
presented in Section \ref{sec:results}.  We point out the differences arising
from the use of two different generators (Wegner and White generator), and
give an overview of the ground state energies up to 42 particles. In
particular, we compare the results obtained with five different \textit{ab
  initio} methods: IM-SRG(2), DMC, Hartree-Fock, FCI and CC. Section
\ref{sec:conclusions} concludes our work and gives perspectives for future
work.

The harmonic oscillator (HO) is a ubiquitous system in quantum physics. It is the basis for an extremely diverse set of many-body calculations -- in nuclear physics \cite{HJORTHJENSEN1995125}, quantum chemistry \cite{helgaker2000molecular}, and quantum dot calculations of solid state physics \cite{reimann2002}. In these calculations, one typically seeks a few of the lowest eigenvalues $E_k$ and eigenvectors of a many-body Hamiltonian $\hat{H}$ that describes the full system by solving the Schr\"odinger equation,
\begin{equation*}
\hat{H} \ket{\psi_k} = E_k \ket{\psi_k}, \quad k \in \{1, 2, \ldots, k_{\text{max}}\}.
\end{equation*}
In all cases, the many-body wave function is expanded in a basis of
eigenstates of the one-body HO system.  Necessarily, in order to perform
numerical calculations the basis must be truncated to give an approximation.
In particular, the so-called ``curse of dimensionality'' implies that for a
system with many particles the number of available degrees of freedom per
particle is severely limited by the finiteness of a computer.

[[This paragraph seems to be talking about a different paper]]
It is clear, that an understanding of the properties of such expansions,
giving \emph{a priori} error estimates on many-body calculations, is very
important. Unfortunately, this is a neglected topic in the physics
literature. In this article, we give a thorough mathematical and numerical
analysis and give practical convergence estimates of HO basis expansions. It
generalizes and refines the findings of a recent study of one-dimensional
systems.\cite{Kvaal2007}

The HO eigenfunctions are popular for several reasons. Many quantum systems, such as the quantum dot model considered here, can be considered perturbed harmonic oscillators \textit{per se}, thus the true eigenstates should be perturbations of the HO eigenstates.  Moreover, the HO has many beautiful properties, such as separability, invariance under orthogonal coordinate changes, and easily computable eigenfunctions, so that computing matrix elements of relevant operators becomes relatively simple.  Since HO eigenfunctions are defined on the \emph{entire} space in which particles live, a truncation of the spatial domain is unnecessary.  This avoids one of the main problems with approaches such as finite difference or finite element methods.\cite{ram2002finite} [[Unknown citation]]

@@

Even though the model system contains 5 parameters $(N, m, \omega, e, \epsilon)$, many of them are redundant.  With an appropriate choice of units, the number of parameters in the system can be reduced to just two.  For rest of this paper, we shall use atomic units\footnote{Here, ``atomic unit'' is used in an \emph{abstract} sense: we do not assume (nor care) whether $m$, $e$, $\epsilon$ correspond to true values in vacuum, or some effective value in medium.} where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ all have a value of one.  Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

The behavior of each electron in isolation is described by the kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2 &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2 &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r})
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{[1]} + \hat H^{[2]}
\end{align*}
where $\hat{H}^{[1]}$ and $\hat{H}^{[2]}$ are the one- and two-body parts of the Hamiltonian,
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}^{[1]} &\equiv \sum_{j = 1}^N h(\hat{\bm p}_j, \hat{\bm r}_j) &
\hat{H}^{[2]} &\equiv \sum_{j_1 = 1}^N \sum_{j_2 = 1}^{j_1 - 1} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align}
The operator $\hat{\bm r}_j$ is the position operator of the $j$-th particle,\footnote{The ordering of the particle labels $j$ is unimportant as they exist only for the purpose of bookkeeping.} and $\hat{\bm p}_j \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_j}$ is its momentum operator.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/fig-shell-structure-v2}
  \caption{Visualization of the 20 lowest single-particle states in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  The $n$ quantum number increases upward vertically, starting at zero for the lowest state within a each column.}
  \label{fig:shell-structure}
\end{figure}

The noninteracting part of the many-body Hamiltonian $\hat{H}^{[1]}$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  The two-dimensional harmonic oscillator Hamiltonian $\hat{h}$ has well known analytic solutions in Cartesian coordinates, but for our purposes, we prefer to use the Fock-Darwin states, which favor a representation in polar coordinates $\bm{r} = (\rho, \varphi)$ and have the useful property of conserving the orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \partial / (\partial \varphi)$.

The Fock-Darwin wave functions can be decomposed into radial and angular components,\cite[Appx.\ A]{lohne2010coupled}
\begin{align}
  F_{n, m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n, |m_\ell|}(\sqrt \omega \rho) A_{m_\ell}(\varphi) \label{eq:fockdarwin} \\
  R_{n, m}(\varrho) &\equiv \sqrt{\frac{2 \cdot n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 / 2} \varrho^m L_n^{(m)}(\varrho^2) \notag \\
  A_{m_\ell}(\varphi) &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i} m_\ell \varphi} \notag
\end{align}
where $L_n^{(m)}$ denotes the generalized Laguerre polynomial of degree $n$ and parameter $m$,
\begin{align*}
  L_n^{(m)}(u) = \frac{1}{n!} u^{-m} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{m + n})
\end{align*}
The states are labeled by two quantum numbers: the principal quantum number $n \in \{0, 1, \ldots\}$ and the orbital angular momentum projection $m_\ell \in \mathbb Z$, which is the eigenvalue of $\hat{L}_{\mathrm{z}}$.

Since electrons are spin-$1/2$ fermions, they can occupy either of the two possible spin states $\chi_{-1/2}$ or $\chi_{+1/2}$.  The complete single-particle basis state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ contains both a spatial component \eqref{eq:fockdarwin} and a spin component,
\begin{align} \label{eq:singleparticlestate}
  \langle (\rho, \varphi) | (n, m_\ell, m_{\mathrm{s}}) \rangle \equiv F_{n, m_\ell}(\rho, \varphi) \chi_{m_{\mathrm{s}}}
\end{align}
where we have introduced the the spin projection $m_{\mathrm{s}} \in \bigl\{-1/2, +1/2\bigr\}$ as the third quantum number.

The energy of the single-particle state $\phi_{n, m_\ell, m_{\mathrm{s}}}$ is given by
\begin{align*}
  \varepsilon_{n, m_\ell, m_{\mathrm{s}}} \equiv (2 n + |m_\ell| + 1) \omega
\end{align*}
The energies are degenerate with respect to the spin projection $m_{\mathrm{s}}$ as our Hamiltonian $\hat{h}$ does not distinguish between them.  Additionally, they are degenerate with respect to the nonnegative integer
\begin{align} \label{eq:shell_index}
  k \equiv 2 n + |m_\ell|
\end{align}
which we conveniently call the \textit{shell index} as it labels each shell starting from zero.  The shells are equidistant with an energy spacing equal to the frequency $\omega$.

Going back to the many-body Hamiltonian $\hat{H}^{[1]}$, we can use the Slater determinant construction to build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the many-body Hamiltonian $\hat{H}^{[1]}$ from an arbitrary choice of $N$ single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$ from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation of the triplet $(n, m_\ell, m_{\mathrm{s}})$.  It is important to emphasize that a Slater determinant $|p_1, \ldots, p_N\rangle$ is not merely a product of the single-particle states, but an antisymmetrized sum over all possible products,
\begin{align*}
  \langle \bm{r}_1, \ldots, \bm{r}_N | p_1, \ldots, p_N \rangle \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|
\end{align*}
The Slater determinant $|p_1, \ldots, p_N\rangle$ is said to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By construction, Slater determinants satisfy the Pauli exclusion principle, ensuring that the exchange of any two particles is equivalent to a reversal in the sign of the many-body wave function.

@@

Currently, the Hamiltonian $\hat{H}$ is explicit in the number of particles $N$.  The many-body machinery can be made agnostic to number of particles through the formalism of second quantization, similar to quantum field theory.  The many-body Hilbert space is extended from states with only $N$ particles to states with every possible number of particles via the Fock space construction.  Slater determinants can then rewritten as,
\begin{align*}
  |p_1, \ldots, p_N\rangle =
  \hat a_{p_1}^\dagger \cdots \hat a_{p_N}^\dagger |\rangle
\end{align*}
whre $|\rangle$ is the \textit{vacuum state} -- the unique state that is devoid of particles -- and $\hat{a_p}^\dagger$ denotes a \textit{creation operator} for the single-particle state $p$.  If the single-particle state $p$ is not yet occupied by a particle, $\hat a_p^\dagger$ adds such a particle to the state.  Otherwise, it collapses the entire many-body state to the number $0$ (not to be confused with the vacuum state $|\rangle$).  The Hermitian conjugate of $\hat a_p^\dagger$ is \textit{annihilation operator} $\hat a_p$, which has the opposite effect of removing the particle if it exists, collapsing to zero otherwise.  For fermions, the field operators $\hat a_p$ and $\hat a_p^\dagger$ satisfy the essential anticommutation relations,
\begin{align*}
  &\{\hat a_p, \hat a_q\} = \{\hat a_p^\dagger, \hat a_q^\dagger\} = 0 &
  &\{\hat a_p, \hat a_q^\dagger\} = \delta_{p, q}
\end{align*}
where $\delta_{p, q}$ denotes the Kronecker delta.

>>>>

\input{head_revtex.tex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{color}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{siunitx}
\usepackage{subcaption}

@@

\begin{document}
\title{Addition and removal energies of circular quantum dots}

\author{Fei Yuan}
\email{yuan@nscl.msu.edu}
\homepage{https://people.nscl.msu.edu/~yuan}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Sam Novario}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Nathan Parzuchowski}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Sarah Reimann}
\affiliation{Department of Chemistry and Center for Theoretical and Computational Chemistry, University of Oslo, N-0316 Oslo, Norway}
\author{Scott Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract}
  We present and compare several many-body methods as applied to two-dimensional quantum dots with circular symmetry.  We calculate the approximate ground state energy using a harmonic oscillator basis optimized by Hartree--Fock (HF) theory and further improve the ground state energy using one of two post-HF methods: in-medium similarity renormalization group (IM-SRG) and coupled cluster singles-and-doubles (CCSD).  With the application of quasidegenerate perturbation theory (QDPT) or the equation-of-motion (EOM) method, we obtain addition and removal energies as well.  Our results are benchmarked against full configuration interaction (FCI) and diffusion Monte Carlo (DMC) where available.  We examine the rate of convergence and perform extrapolations to the infinite basis limit using a power-law model.
\end{abstract}

@@

The behavior of strongly confined electrons is of fundamental concern to many-body theory.  The prototypical system is that of quantum dots, also known as ``artificial atoms'', in which electrons are confined within artificial constructed semiconducting heterostructures.  Such nanoscale systems are highly relevant as they lie at a critical point where both theoretical models and experimental observations can readily probe quantum phenomena such as tunneling, entanglement, and magnetization \cite{reimann2002,engel1993}.  Moreover, unlike many physical systems, quantum dots enjoy the benefit of being highly tunable through changes in the external field or the structure of the confining material.  This allows experiments to quantify the impact of quantum effects at different levels of correlation.

The ground states of quantum dots exhibit shell structures and magic numbers similar to those of atoms and nuclei \cite{tarucha1996}.  Thus, this provides a rare opportunity to study electronic systems without the influence of the atomic nuclei.  They also provide a testbed for the development of many-body methods for other systems with similar geometry, such as nuclei or neutrons drops.

Beyond their theoretical relevance, there are numerous potential applications quantum dots.  The electrical and optical properties of quantum dots are particularly useful for solar cells \cite{jenks:013111,doi:10.1021/cr900289f} and laser technology \cite{strauf2010,5075760}.  They are have potential medical applications in diagnosis and treatment \cite{Ben-Ari02042003}.  And of course, quantum dots are also promising candidates for the physical realization of quantum computing\cite{PhysRevA.57.120}.

In general, quantum dot systems are not analytically solvable, with the exception of 2-electron systems or systems with specific strengths of the external field \cite{PhysRevA.48.3561}.  Hence, in practical applications one can only hope for a numerical solution through some \textit{many-body method} or a combination thereof.

One of the most accurate techniques for solving many-body systems is that of exact diagonalization, also known as the full configuation interaction method.  Unlike most methods, it is assured to converge to the exact answer as the size of the finite basis is increases to infinity (the so-called \textit{infinite-basis limit}).  Despite this significant advantage, it is often infeasible to perform exact diagonalization due to its factorially increasing cost with respect to the basis size and the number of particles.  This led to the development of more cost-effective methods that trade varying amounts of accuracy for varying amounts of speed.

In this paper, we analyze the application of several methods to the quantum dot systems, including the Hartree--Fock (HF) method, M\o ller-Plesset perturbation theory (MPPT), the in-medium similarity renormalization group (IM-SRG) method, coupled cluster (CC) theory \cite{PhysRevB.67.045320,heidari:114708,PhysRevB.84.115302}, quasidegenerate perturbation theory (QDPT), and equation-of-motion (EOM) methods.  We compare them against variational and diffusion Monte Carlo (VMC and DMC) \cite{PhysRevB.68.035304,PhysRevB.62.8120,PhysRevB.84.115302,PhysRevB.54.4780} and full configuration interaction (FCI) \cite{olsen2013thesis,JJAP.36.3924,PhysRevB.56.6428,2008arXiv0810.2644K,rontani:124102} results from existing literature.

Of these methods, IM-SRG is a recently developed technique that has shown significant promises.  Similarity renormalization group (SRG) methods are a family of methods that transform the Hamiltonian into a band- or block-diagonal form through a continuous sequence of unitary transformations.  The goal of such a transformation is to reduce the coupling between a subspace of interest -- such as the ground state or a set of low-lying states -- and the remaining Hilbert space.  It has successfully been applied to systems with various underlying potentials to calculate their binding energy and other observables, especially in nuclear theory \cite{ScottSRG,PhysRevC.75.061001,SRGThreeDim}.  The \emph{in-medium} SRG (IM-SRG) is an adaptation of the SRG method to evolve Hamiltonians in a truncated Fock space, which avoids the prohibitive cost of constructing a full many-body Hamiltonian.  The effect of the truncation is minimized through the normal-ordering technique, which re-expands the operators with respect to a nearby many-body reference state rather than the far-away vacuum state.

The article is organized as follows: Section \ref{sec:formalism} introduces first (\ref{subsec:modelHamiltonian}) the Hamiltonian we use to model circular quantum dots, and gives afterwards an overview of the \textit{ab initio} many-body methods used in this paper: Hartree--Fock, IM-SRG, CCSD, QDPT, EOM.  Our results are presented in Section \ref{sec:results}.  We analyze and compare the differences between the various methods and also between different quantum dot systems.  We examine the usefulness extrapolation techniques in improving the precision of the results.  Section \ref{sec:conclusions} concludes our work and gives perspectives for future work.

@@

Even though the model system contains 5 parameters $(N, m, \omega, e, \epsilon)$, many of them are redundant.  With an appropriate choice of units, the number of parameters in the system can be reduced to just two.  For rest of this paper, we shall use atomic units\footnote{Here, ``atomic unit'' is used in an \emph{abstract} sense: we do not assume (nor care) whether $m$, $e$, $\epsilon$ correspond to true values in vacuum, or some effective value in medium.} where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ are all set to one.  Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

The behavior of each electron in isolation is described by the nonrelativistic kinetic energy $t(\bm{r})$ and the potential energy due to the trap $u(\bm{r})$, which combine to form the standard harmonic oscillator Hamiltonian $h(\bm{p}, \bm{r})$:
\begin{align*}
  t(\bm{p}) &\equiv \frac{1}{2} \bm{p}^2 &
  u(\bm{r}) &\equiv \frac{1}{2} \omega^2 \bm{r}^2 &
  h(\bm{p}, \bm{r}) &\equiv t(\bm{p}) + t(\bm{r})
\end{align*}
Here, $\bm{p}$ is the linear momentum of the electron and $\bm{r}$ is its position relative to the center of the trap.  The pairwise interaction between two electrons at positions $\bm{r}$ and $\bm{r}'$ is simply the Coulomb interaction:
\begin{align*}
  v(\bm{r}, \bm{r}') \equiv \frac{1}{|\bm{r} - \bm{r}'|}
\end{align*}
Then, the full many-body problem is described by the Hamiltonian
\begin{align*}
  \hat H \equiv \hat H^{[1]} + \hat H^{[2]}
\end{align*}
where $\hat{H}^{[1]}$ and $\hat{H}^{[2]}$ are the one- and two-body parts of the Hamiltonian,
\begin{align} \label{eq:onetwobodyhamiltonian}
\hat{H}^{[1]} &\equiv \sum_{j = 1}^N h(\hat{\bm p}_j, \hat{\bm r}_j) &
\hat{H}^{[2]} &\equiv \sum_{j_1 = 1}^N \sum_{j_2 = 1}^{j_1 - 1} v(\hat{\bm r}_{j_1}, \hat{\bm r}_{j_2})
\end{align}
The operator $\hat{\bm r}_j$ is the position operator of the $j$-th particle,\footnote{The ordering of the particle labels $j$ is unimportant as they exist only for the purpose of bookkeeping.} and $\hat{\bm p}_j \equiv -\mathrm{i} \hat{\bm{\nabla}}_{\bm{r}_j}$ is its momentum operator.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/fig-shell-structure-v2}
  \caption{Visualization of the 20 lowest single-particle states in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  The $n$ quantum number increases upward vertically, starting at zero for the lowest state within each column.}
  \label{fig:shell-structure}
\end{figure}

The noninteracting part of the many-body Hamiltonian $\hat{H}^{[1]}$ can be reduced to $N$ independent single-particle Hamiltonians each of the form:
\begin{align*}
  \hat{h} = \frac{1}{2} \hat{\bm{\nabla}}^2 + \frac{1}{2} \omega^2 \hat{\bm{r}}^2
\end{align*}
Therefore, solving the noninteracting many-body problem amounts to solving the single-particle Schr\"odinger equation
\begin{align*}
  \hat{h} |p\rangle = \varepsilon_p |p\rangle
\end{align*}
where $p$ is an unknown tuple of quantum numbers.  In Cartesian coordinates, the two-dimensional harmonic oscillator is trivially reducible to the well-known one-dimensional problem.  However, to exploit the circular symmetry, we prefer to use Fock-Darwin states, which are written in polar coordinates $\bm{r} = (\rho, \varphi)$.  Such states have the computationally useful property of conserving orbital angular momentum $\hat{L}_{\mathrm{z}} \equiv -\mathrm{i} \partial / (\partial \varphi)$.

The Fock-Darwin wave functions can be decomposed into radial and angular components,\cite[Appx.\ A]{lohne2010coupled}
\begin{align}
  F_{n, m_\ell}(\rho, \varphi) &\equiv \sqrt\omega R_{n, |m_\ell|}(\sqrt \omega \rho) A_{m_\ell}(\varphi) \label{eq:fockdarwin} \\
  R_{n, m}(\varrho) &\equiv \sqrt{\frac{2 \cdot n!}{(n + m)!}} \mathrm{e}^{-\varrho^2 / 2} \varrho^m L_n^{(m)}(\varrho^2) \notag \\
  A_{m_\ell}(\varphi) &\equiv \frac{1}{\sqrt{2 \pi}} \mathrm{e}^{\mathrm{i} m_\ell \varphi} \notag
\end{align}
where $L_n^{(\alpha)}$ denotes the (generalized) Laguerre polynomial \cite[\S 18.3]{NIST:DLMF} of degree $n$ and parameter $\alpha$,
\begin{align*}
  L_n^{(\alpha)}(u) = \frac{1}{n!} u^{-\alpha} \mathrm{e}^u \frac{\mathrm{d}^n}{\mathrm{d} u^n} (\mathrm{e}^{-u} u^{\alpha + n})
\end{align*}
The states are labeled by two quantum numbers: the principal quantum number $n \in \{0, 1, 2, \ldots\}$ and orbital angular momentum projection $m_\ell \in \mathbb \{\ldots, 2, -1, 0, 1, 2, \ldots\}$.  For a wave function, $n$ indicates the degree of the Laguerre polynomial, whereas $m_\ell$ is simply the eigenvalue of $\hat{L}_{\mathrm{z}}$.

Since electrons are spin-$1/2$ fermions, they can occupy either of the two possible spin states $\chi_{-1/2}$ or $\chi_{+1/2}$.  Thus, every single-particle basis state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ contains both a spatial component \eqref{eq:fockdarwin} and a spin component,
\begin{align} \label{eq:singleparticlestate}
  \langle (\rho, \varphi), m_{\mathrm{s}}' | (n, m_\ell, m_{\mathrm{s}}) \rangle \equiv F_{n, m_\ell}(\rho, \varphi) \delta_{m_{\mathrm{s}}^{}, m_{\mathrm{s}}'}
\end{align}
Here we have introduced spin projection $m_{\mathrm{s}} \in \bigl\{-1/2, +1/2\bigr\}$ as the third quantum number and $\delta_{p, q}$ to denote the Kronecker delta..

The energy of the single-particle state $| (n, m_\ell, m_{\mathrm{s}}) \rangle$ is given by
\begin{align} \label{eq:energysingleparticlestate}
  \varepsilon_{n, m_\ell, m_{\mathrm{s}}} \equiv (2 n + |m_\ell| + 1) \omega
\end{align}
The energies are degenerate with respect to the spin projection $m_{\mathrm{s}}$ as our Hamiltonian $\hat{h}$ does not distinguish between them.  Additionally, they are degenerate with respect to the nonnegative integer
\begin{align} \label{eq:shell_index}
  k \equiv 2 n + |m_\ell|
\end{align}
which we conveniently call the \textit{shell index} as it labels each shell starting from zero.  The shells are equidistant with an energy spacing equal to the frequency $\omega$.

Returning to the many-body Hamiltonian $\hat{H}^{[1]}$, we can use the Slater determinant construction to build an $N$-particle eigenstate $|p_1, \ldots, p_N\rangle$ of the many-body Hamiltonian $\hat{H}^{[1]}$ from an arbitrary choice of $N$ single-particle eigenstates $|p_1\rangle, \ldots, |p_N\rangle$ from \eqref{eq:singleparticlestate}, where $p$ is an abbreviation for the triplet $(n, m_\ell, m_{\mathrm{s}})$.  Note that a Slater determinant $|p_1, \ldots, p_N\rangle$ is not merely a product of the single-particle states, but an antisymmetrized sum over all possible products,
\begin{align*}
  \langle \bm{r}_1, \ldots, \bm{r}_N | p_1, \ldots, p_N \rangle \equiv
  \frac{1}{\sqrt{N}} \left|
  \begin{matrix}
    \langle \bm{r}_1 | p_1 \rangle & \cdots & \langle \bm{r}_1 | p_N \rangle \\
    \vdots & \ddots & \vdots \\
    \langle \bm{r}_N | p_1 \rangle & \cdots & \langle \bm{r}_N | p_N \rangle \\
  \end{matrix}
  \right|
\end{align*}
The Slater determinant $|p_1, \ldots, p_N\rangle$ is said to \textit{occupy} the states $|p_1\rangle, \ldots, |p_N\rangle$.  By construction, Slater determinants satisfy the Pauli exclusion principle, ensuring that the exchange of any two particles is equivalent to a reversal in the sign of the many-body wave function.

@@

Currently, the Hamiltonian $\hat{H}$ is defined for a fixed number of particles $N$.  Through the formalism of second quantization, the many-body machinery can be made agnostic to number of particles, in analogy to quantum field theory.  Instead of considering the Hilbert space for exactly $N$ particles, we combine every space with $N = 0, 1, 2, \ldots$ particles into a combined Hilbert space called the Fock space.  To establish relationships between states with differing number of particles, we introduce the \textit{field operators} $\hat{a}_p^\dagger$ and $\hat{a}_p$ to increment and decrement the number of particles in a state, respectively.  The linear operator $\hat{a}_p^\dagger$ is known as \textit{creation operator} for the single-particle state $p$.  Its action on an arbitrary Slater determinant is defined as:
\begin{align*}
  \hat a_p^\dagger |p'_1, \ldots, p'_N\rangle =
  |p, p'_1, \ldots, p'_N\rangle
\end{align*}
Informally, it is said to \textit{create} a particle in the state $p$.  If $p$ coincides with any of the existing $p'_1, \ldots, p'_N$, then antisymmetrization would collapse the state to the number $0$, an invalid state.  The Hermitian conjugate of $\hat{a}_p^\dagger$ is $\hat{a}_p$, which known as the \textit{annihilation operator} for the state $p$.  By analogy, its action is to remove the state $p$ from a Slater determinant, or collapse it to the number $0$ if the state $p$ was not occupied in the Slater determinant.
